---
title: "Negative Equity Trends in the U.S. Housing Market"
author: "Sean Mussenden | Howard Center for Investigative Journalism"
date: "11/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This memo describes initial data analysis done to identify clusters in the United States communities where a significant portion of homes are "underwater", a decade after the sharp decline in home prices during the global financial crisis. An underwater home is one where the estimated value of a home is lower than the estimated principal balance of the mortgage (negative loan to value or LTV). The analysis also seeks to identify demographic and economic features of communities with high rates of negative LTV homes, and to understand how negative LTV rates have changed since 2009, when some communities had a majority of homes with negative LTVs. 

# Major findings

* Rates today are nowhere near where they were in 2009. 
* Some geographic concentration today. 
* 

# Load Packages, Define Functions, Change Settings ###

```{r include=FALSE}

# Turn off scientific notation
options(scipen=9999)

## Load packages
# For general data science goodness
library(tidyverse)

# For data cleaning
library(janitor)

# For working with datetime
library(lubridate)

# For reading in Excel files
library(readxl)

# For working with ZIP Codes
library(zipcode)

# For mapping
library(maps)
library(mapview)
library(sf)
library(leaflet)
library(leafpop)
library(leafem)
library(raster)
library(tigris)

# For pulling census data
library(tidycensus)

# For correlations
library(corrr)
library(moderndive)
library(Hmisc)
library(broom)
# For graphics
library(scales)
library(ggthemes)
library(DT)
library(ggpubr)

# Function to flatten correlation matrix
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}

```


# Data description, loading and cleaning

We used several data sets describing underwater rates, real estate metrics, and community demographic and income factors in this analysis. 

## Underwater rates 

We have two data negative equity via the real estate analytics firm CoreLogic: Negative equity share of all homes by county by month-year from 2009 to present and negative equity share of all homes by ZIP code in 2019.  Note: We are waiting for a third data set, which they sent to us, but had errors I identified (only had data for four states), of negative equity share of all homes by ZIP code by month-year from 2009-present.

```{r include=FALSE}

# Load negative equity share by county by month-year
underwater_county_month_year <- read_xlsx("../data/input_data/core_logic/negative_equity_share_county.xlsx")

# Clean negative equity share by county by month-year
underwater_county_month_year <- underwater_county_month_year %>%
  # create year column from yyyymm
  mutate(year=str_sub(yyyymm, 1,4)) %>%
  # create month column yyyymm
  mutate(month=str_sub(yyyymm,5,6)) %>%
  # select needed columns
  dplyr::select(state_code, fips_code, state_name, county_name,year, month, yyyymm, percent_negative_equity) %>%
  # fix busted fips codes by converting to character and adding a leading 0 to four digit codes
  mutate(fips_code = as.character(fips_code)) %>%
  mutate(fips_code = case_when(str_length(fips_code) < 5  ~ paste0("0",fips_code),
                                TRUE ~ fips_code)
         ) %>%
  mutate(percent_negative_equity = round(percent_negative_equity*100, 2)) 

# Create negative equity share by county by year, using average of 12 months in a given year as value for year.
underwater_county_year <- underwater_county_month_year %>%
  # Group by county and year
  group_by(state_code, fips_code, state_name, county_name, year) %>%
  # Average 12 months in each year
  summarise(percent_negative_equity = round(mean(percent_negative_equity),2)) %>%
  # Make the long data wide
  spread(year, percent_negative_equity) %>%
  # Fix column names
  rename_at(vars(matches("20")), funs(paste0("pct_negative_equity_y", .)))

# Remove underwater_county_month_year
rm(underwater_county_month_year)

# Load and clean ZIP Code from  2019
# Still need to get from CoreLogic an answer on time period

underwater_zips_2019 <- read_xlsx("../data/input_data/core_logic/corelogic_underwater_homes.xlsx") 

# Clean ZIP code from 2019
underwater_zips_2019 <- underwater_zips_2019 %>%
  # Fix column names
  clean_names() %>%
  # Fix zip codes
  mutate(zip_code = clean.zipcodes(zip_code)) %>%
  # Make negative equity share readable and standardize column name
  mutate(percent_negative_equity = round(share_of_homes_in_negative_equity*100, 2)) %>%
  dplyr::select(-share_of_homes_in_negative_equity)

### Load and clean ZIP Code for every month in dating back a decade
### Can't use this because it after doing some initial analysis and cleaning, it appears to only have data from five states: Alabama, Alaska, Idaho, Hawaii and Georgia. state fips codes 01, 02, 13, 15 and 16.
# underwater_zips_all_months <- read_csv("../data/input_data/core_logic/zip_negative_equity_extended.csv") 

```

## Demographic and Economic Data

We pulled select racial, ethnic origin and economic variables from the U.S. Census American Community Survey, via the Tidycensus package.

``` {r include=FALSE}

# Load variables for exploration
# acs_variable <- load_variables(2017, "acs5", cache = TRUE)

# Define census api key
census_api_key("549950d36c22ff16455fe196bbbd01d63cfbe6cf")

# Get percent white, percent black, percent hispanic, poverty rate, income data by county

acs_county_white <- get_acs(geography = "county", variables = c("B02001_002"), geometry = FALSE, survey="acs5", year = 2017, summary_var ="B01001_001") %>%
  mutate(pct_white_2017 = estimate/summary_est) %>%
  dplyr::select(GEOID,NAME,pct_white_2017) %>%
  clean_names()

acs_county_black <- get_acs(geography = "county", variables = c("B02001_003"), geometry = FALSE, survey="acs5", year = 2017, summary_var ="B01001_001") %>%
  mutate(pct_black_2017 = estimate/summary_est) %>%
  dplyr::select(GEOID,NAME,pct_black_2017) %>%
  clean_names()

acs_county_hispanic <- get_acs(geography = "county", variables = c("B03001_003"), geometry = FALSE, survey="acs5", year = 2017, summary_var ="B01001_001") %>%
  mutate(pct_hispanic_2017 = estimate/summary_est) %>%
  dplyr::select(GEOID,NAME,pct_hispanic_2017) %>%
  clean_names()

acs_county_poverty <- get_acs(geography = "county", variables = c("B06012_002"), geometry = FALSE, survey="acs5", year = 2017, summary_var ="B01001_001") %>%
  mutate(pct_poverty_2017 = estimate/summary_est) %>%
  dplyr::select(GEOID,NAME,pct_poverty_2017) %>%
  clean_names()

acs_county_income <- get_acs(geography = "county", variables = c("B19013_001"), geometry = FALSE, survey="acs5", year = 2017) %>%
  dplyr::select(GEOID,NAME, median_household_income_2017 = estimate) %>%
  clean_names()

# Join the variables into a single data frame

acs_data_county <- acs_county_white %>%
  inner_join(acs_county_black) %>%
  inner_join(acs_county_hispanic) %>%
  inner_join(acs_county_poverty) %>%
  inner_join(acs_county_income) %>%
  mutate(pct_white_2017 = round(pct_white_2017*100,2),
         pct_black_2017 = round(pct_black_2017*100,2),
         pct_hispanic_2017 = round(pct_hispanic_2017*100,2),
         pct_poverty_2017 = round(pct_poverty_2017*100,2),
         ) 
  
  
# Get list of fipscodes from Tigris package
fips_codes <- fips_codes %>%
  mutate(fips_code = paste0(state_code,county_code)) 

# Join fips codes to table of census variables
acs_data_county <- acs_data_county %>%
  inner_join(fips_codes, by=c("geoid" = "fips_code"))

# Remove everything except for single acs_data_county table
rm(list=ls(pattern="acs_county"))

# Get percent white, percent black, percent hispanic, poverty rate, income data by ZCTA

acs_zcta_white <- get_acs(geography = "zcta", variables = c("B02001_002"), geometry = FALSE, survey="acs5", year = 2017, summary_var ="B01001_001") %>%
  mutate(pct_white_2017 = estimate/summary_est) %>%
  dplyr::select(GEOID,NAME,pct_white_2017) %>%
  clean_names()

acs_zcta_black <- get_acs(geography = "zcta", variables = c("B02001_003"), geometry = FALSE, survey="acs5", year = 2017, summary_var ="B01001_001") %>%
  mutate(pct_black_2017 = estimate/summary_est) %>%
  dplyr::select(GEOID,NAME,pct_black_2017) %>%
  clean_names()

acs_zcta_hispanic <- get_acs(geography = "zcta", variables = c("B03001_003"), geometry = FALSE, survey="acs5", year = 2017, summary_var ="B01001_001") %>%
  mutate(pct_hispanic_2017 = estimate/summary_est) %>%
  dplyr::select(GEOID,NAME,pct_hispanic_2017) %>%
  clean_names()

acs_zcta_poverty <- get_acs(geography = "zcta", variables = c("B06012_002"), geometry = FALSE, survey="acs5", year = 2017, summary_var ="B01001_001") %>%
  mutate(pct_poverty_2017 = estimate/summary_est) %>%
  dplyr::select(GEOID,NAME,pct_poverty_2017) %>%
  clean_names()

acs_zcta_income <- get_acs(geography = "zcta", variables = c("B19013_001"), geometry = FALSE, survey="acs5", year = 2017) %>%
  dplyr::select(GEOID,NAME, median_household_income_2017 = estimate) %>%
  clean_names()

# Join the variables into a single data frame

acs_data_zcta <- acs_zcta_white %>%
  inner_join(acs_zcta_black) %>%
  inner_join(acs_zcta_hispanic) %>%
  inner_join(acs_zcta_poverty) %>%
  inner_join(acs_zcta_income) %>%
  mutate(pct_white_2017 = round(pct_white_2017*100,2),
         pct_black_2017 = round(pct_black_2017*100,2),
         pct_hispanic_2017 = round(pct_hispanic_2017*100,2),
         pct_poverty_2017 = round(pct_poverty_2017*100,2),
         )
  

# Remove everything except for single acs_data_zcta table

rm(list=ls(pattern="acs_zcta|fips_codes"))

```

## Load and clean Zillow data

From the real estate data firm Zillow, we pulled information on current, historical and forecast community home values. Zillow has also done extensive work on underwater homes, but stopped releasing data in 2017. 

```{r include=FALSE}

# https://www.zillow.com/research/data/

# Read in Zillow Region ID to County Crosswalk
zillow_crosswalk <- read_csv("../data/input_data/zillow/CountyCrossWalk_Zillow.csv") %>%
  mutate(fips_code = as.character(FIPS)) %>%
  mutate(fips_code = case_when(str_length(fips_code) < 5  ~ paste0("0",fips_code),
                                TRUE ~ fips_code)
         ) %>%
  dplyr::select(CountyRegionID_Zillow, fips_code) 

# Home values summary for current month (Oct 2019) and historical comparison by county 

county_ZHVI_summary_current_month <- read_csv("../data/input_data/zillow/County_Zhvi_Summary_AllHomes.csv") %>%
  mutate(five_year_pct_change = round(`5Year`*100,2),
         ten_year_pct_change = round(`10Year`*100,2),
         month_pct_change = round(MoM*100,2),
         quarter_pct_change = round(QoQ*100,2),
         year_pct_change = round(YoY*100,2)) %>%
  inner_join(zillow_crosswalk, by=c("RegionID" = "CountyRegionID_Zillow")) %>%
  dplyr::select(fips_code, RegionName, State, Zhvi, month_pct_change, quarter_pct_change, year_pct_change, five_year_pct_change, ten_year_pct_change)


# Home values summary for current month (Oct 2019) and historical comparison by ZIP

Zip_ZHVI_summary_current_month <- read_csv("../data/input_data/zillow/Zip_Zhvi_Summary_AllHomes.csv") %>%
   mutate(five_year_pct_change = round(`5Year`*100,2),
         ten_year_pct_change = round(`10Year`*100,2),
         month_pct_change = round(MoM*100,2),
         quarter_pct_change = round(QoQ*100,2),
         year_pct_change = round(YoY*100,2)) %>%
  dplyr::select(zip_code = RegionName, State, Zhvi,month_pct_change, quarter_pct_change, year_pct_change, five_year_pct_change, ten_year_pct_change)

# Year over year ZHVI forecast current month (Oct 2019), all geographics (long sheet, sted wide)
all_regions_forecasts <- read_csv("../data/input_data/zillow/AllRegionsForePublic-1.csv")

# Create year over year ZHVI forecast just for counties
county_forecasts <- all_regions_forecasts %>%
  filter(Region == "County")

# Create year over year ZHVI forecast just for ZIP
zip_forecasts <- all_regions_forecasts %>%
  filter(Region=="Zip")

# Join county summary plus forecast
county_summary_forecast <- county_ZHVI_summary_current_month %>%
  inner_join(county_forecasts, by=c("State" = "StateName","RegionName" = "CountyName")) %>%
  mutate(forecast_year_pct_change = round(ForecastYoYPctChange,2)) %>%
  dplyr::select(-Region, -RegionName.y, -CityName, -ForecastYoYPctChange) %>%
  rename(zhvi = Zhvi,
         zhvi_month_pct_change = month_pct_change,
         zhvi_quarter_pct_change = quarter_pct_change,
         zhvi_year_pct_change = year_pct_change,
         zhvi_five_year_pct_change = five_year_pct_change,
         zhvi_ten_year_pct_change = ten_year_pct_change,
         zhvi_forecast_year_pct_change = forecast_year_pct_change
         )


# Join ZIP summary plus forecast
zip_summary_forecast <- Zip_ZHVI_summary_current_month %>%
  inner_join(zip_forecasts, by=c("zip_code" = "RegionName")) %>%
  mutate(forecast_year_pct_change = round(ForecastYoYPctChange,2)) %>%
  dplyr::select(-Region, -StateName, -CountyName, -CityName, -ForecastYoYPctChange) %>%
  rename(zhvi = Zhvi,
         zhvi_month_pct_change = month_pct_change,
         zhvi_quarter_pct_change = quarter_pct_change,
         zhvi_year_pct_change = year_pct_change,
         zhvi_five_year_pct_change = five_year_pct_change,
         zhvi_ten_year_pct_change = ten_year_pct_change,
         zhvi_forecast_year_pct_change = forecast_year_pct_change
         )

# For later analysis Time series home values by County and Zip
#county_ZHVI_time_series <- read_csv("../data/input_data/zillow/County_Zhvi_AllHomes.csv")
#Zip_ZHVI_time_series <- read_csv("../data/input_data/zillow/Zip_Zhvi_AllHomes.csv")

# Remove unneeded files
rm(list=c("county_ZHVI_summary_current_month","Zip_ZHVI_summary_current_month", "all_region_forecasts","zip_forecasts", "county_forecasts", "zillow_crosswalk"))

```

## Load and clean urban-rural county designation and unemployment data

The USDA classifies county on a scale from 1 (most urban) to 9 (most rural).  Unemployment via BLS in 2018.

```{r}
# County rural urban code designation from USDA
# Spectrum from 1 (most urban) to 9 (most rural)
rural_urban <- read_xls("../data/input_data/rural_urban_codes/ruralurbancodes2013.xls") %>%
  rename(fips_code = FIPS)

# Unemployment by county from BLS 2018
unemployment <- read_xlsx("../data/input_data/bls/laucnty18.xlsx") %>%
  mutate(fips_code = paste0(state_fips,county_fips)) %>%
  dplyr::select(fips_code, county_state, unemployment_rate_2018 = unemployment_rate)
  
```

## Join data frames

Put together CoreLogic county/ZIP code level underwater data with U.S. census data, USDA ruralness data, Zillow data.

```{r}


## Make county data frames (2009-2019 time series)

# Underwater data, census data, rural data - 2446 counties
underwater_county_year_no_zillow <- underwater_county_year %>%
  ungroup() %>%
  inner_join(rural_urban) %>%
  dplyr::select(-State,-Population_2010, -County_Name, -Description) %>%
  inner_join(acs_data_county, by=c("fips_code" = "geoid")) %>%
  dplyr::select(fips_code, state_name = state_name.x, county_name, starts_with("RUCC"), starts_with("pct"), starts_with("median")) %>%
  inner_join(unemployment) %>%
  dplyr::select(-county_state)

# Underwater data, census data, rural data, Zillow data - 1477 counties 
underwater_county_year_yes_zillow <- underwater_county_year_no_zillow %>%
  inner_join(county_summary_forecast) %>%
  dplyr::select(-RegionName, -State) 


## Make ZIP Code dataframes (2019 only)

# Underwater data, census data - 28989 zip codes
underwater_zips_2019_no_zillow <- underwater_zips_2019 %>%
  ungroup() %>%
  inner_join(acs_data_zcta, by=c("zip_code" = "geoid")) %>%
  dplyr::select(zip_code, everything(),-name) 

# Underwater data, census data, Zillow data - 13729 zip codes

underwater_zips_2019_yes_zillow <- underwater_zips_no_zillow %>%
  inner_join(zip_summary_forecast) %>%
  dplyr::select(-State) 

rm(list=setdiff(ls(), c("underwater_zips_2019_no_zillow", "underwater_zips_2019_yes_zillow", "underwater_county_year_no_zillow", "underwater_county_year_yes_zillow", "flattenCorrMatrix")))

```
## Load Shapefiles

```{r}
#ZCTAs can take several minutes to download.  To cache the data and avoid re-downloading in future R sessions, set `options(tigris_use_cache = TRUE)`
options(tigris_use_cache = TRUE)

# ZIP Code Points
data(zipcode)

# ZCTA shapefiles
zctas <- zctas(cb=TRUE)

# Counties
counties <- counties(cb = TRUE)


```

# Points to Stress

## Negative Equity Clusters in 2019 

Though negative equity rates are nowhere near 2009 levels, the analysis identified 10 clusters with a higher negative equity rates relative to the rest of the country. 

In the map below, click on the two-letter buttons at right to zoom to that cluster. 

Click on each ZIP code to see info about each county.

Clusters
* MD1 | Maryland+DC | Cluster of ZIP Codes in Prince George's County and Anacostia in D.C. and down through Waldorm, in Charles County.
* MD2 | Maryland Eastern Shore | Huge chunks of the lower Eastern Shore, through Salisbury and Ocean City.
* NJ1 | South New Jersey | Large swaths of southern New Jersey, including Atlantic City, Philly suburbs.
* NJ2 | North Jersey+NYC | New York City suburbs, including Newark, Elizabeth, Paterson, parts of Queens
* CT | Connecticut | Several areas, including Hartford and Waterbury, Bridgeport and New Haven, up to Rhode Island border.
* IA | Iowa | a bunch of communities in Iowa with no real clusters. Can't make heads or tails of this.
* FL | Florida | Worst issues in Miami, Hialeah and Homestead, scatterd parts of the state.
* IL | Chicago | Huge issues surrounding Chicago, especially on South Side
* CA | California | Issues in Fresno, south of Monterey, scattered throughout. 
* NV | Las Vegas | 
* No buttons, but interesting | Baton Rouge, New Orleans, Atlanta, North Dakota


```{r}

# Filter for only high negative equity zip codes
high_negative_equity_zips <- underwater_zips_2019_no_zillow %>%
  filter(percent_negative_equity >= 5)

# Join zip code coordinates to negative equity by zipcode
high_negative_equity_zips_geo <- geo_join(zctas, high_negative_equity_zips, 'GEOID10', 'zip_code', 
how = "inner")

# Color Scheme
binpal <- colorBin("plasma", high_negative_equity_zips_geo$percent_negative_equity, 5, pretty = FALSE)

# Draw map
leaflet(high_negative_equity_zips_geo) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%
   #addProviderTiles(providers$Wikimedia) %>%
   addPolygons(fillColor = ~binpal(percent_negative_equity), weight = 1, smoothFactor = 0.5, opacity = 0.1, fillOpacity = 0.5, color="black", popup = popupTable(high_negative_equity_zips_geo)) %>%
   setView(-95, 39.335359608681216, 4) %>%
   addHomeButton(extent(-98.39355468750001, -89.60449218750001, 39.85072092501597, 43.8899753738369), "IA") %>%
   addHomeButton(extent(-115.44021606445314, -114.89089965820314, 36.04521273039952, 36.318998009207924), "NV") %>%
   addHomeButton(extent(-127.24365234375001, -109.66552734375001, 31.98944183792288, 40.713955826286046), "CA") %>%
   addHomeButton(extent(-88.33145141601564, -87.23281860351564, 41.466399253078876, 41.97276436226528), "IL") %>%
   addHomeButton(extent(-80.82092285156251, -79.72229003906251, 25.517657429994035, 26.12831612064242), "FL") %>%
   addHomeButton(extent(-73.95446777343751, -71.75720214843751, 41.03793062246529, 42.05337156043361), "CT") %>%
   addHomeButton(extent(-74.46807861328126, -73.91876220703126, 40.57484790030712, 40.83199550584334), "NJ2") %>%
   addHomeButton(extent(-75.86883544921876, -73.67156982421876, 39.16414104768742, 40.20824570152502), "NJ1") %>%
   addHomeButton(extent(-76.95098876953126, -74.75372314453126, 38.10214399750345, 39.16201148082406), "MD2") %>%
   addHomeButton(extent(-77.1947479248047, -76.6454315185547, 38.791021386961596, 39.0549177529185), "MD1") %>%
   addHomeButton(extent(-131.74804687500003, -61.43554687500001, 18.812717856407776, 52.908902047770255 ), "U.S.") %>%
   
   addLegend("bottomleft", 
             pal = colorBin("plasma", high_negative_equity_zips_geo$percent_negative_equity), 
             values = high_negative_equity_zips_geo$percent_negative_equity,
    title = "% Homes Negative Equity 2019",
    labFormat = labelFormat(prefix = ""),
    opacity = 1
  ) %>%
  addEasyButton(easyButton(
    states = list(
      easyButtonState(
        stateName="unfrozen-markers",
        icon="ion-toggle",
        title="Get Bounding box",
        onClick = JS("
                     function(btn, map) {
                        alert(
                         map.getBounds().getWest() + ', ' + map.getBounds().getEast() + ', ' + map.getBounds().getSouth() + ', ' + map.getBounds().getNorth() 
                        );
                        
                     }")
      )
    )
  )
)

```

```{r}

datatable(high_negative_equity_zips, class = 'cell-border stripe', filter = 'top', rownames = FALSE, caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    '2019: ZIP codes with negative equity >= 4%'
  ),
  options = list(scrollX=TRUE, scrollCollapse=TRUE)
)
```

This map shows county-level averages. There were 89 counties with a negative equity rate above 4 percent in 2019.

```{r}

# Filter for 2019
high_negative_equity_county_2019 <- underwater_county_year_no_zillow %>%
  #dplyr::select(state_code, fips_code, state_name, county_name, y2019) %>%
  filter(pct_negative_equity_y2019 >= 4) %>%
  arrange(desc(pct_negative_equity_y2019))

# Join zip code coordinates to negative equity by zipcode
high_negative_equity_county_2019_geo <- geo_join(counties, high_negative_equity_county_2019, 'GEOID', 'fips_code', 
how = "inner")  

# Color Scheme
binpal <- colorBin("plasma", high_negative_equity_county_2019$pct_negative_equity_y2019, 5, pretty = FALSE)

# Draw map
leaflet(high_negative_equity_county_2019_geo) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%
   #addProviderTiles(providers$Wikimedia) %>%
   addPolygons(fillColor = ~binpal(pct_negative_equity_y2019), weight = 1, smoothFactor = 0.5, opacity = 0.1, fillOpacity = 0.5, color="black", popup = popupTable(high_negative_equity_county_2019_geo)) %>%
   setView(-95, 39.335359608681216, 4) %>%
   addHomeButton(extent(-98.39355468750001, -89.60449218750001, 39.85072092501597, 43.8899753738369), "IA") %>%
   addHomeButton(extent(-115.44021606445314, -114.89089965820314, 36.04521273039952, 36.318998009207924), "NV") %>%
   addHomeButton(extent(-127.24365234375001, -109.66552734375001, 31.98944183792288, 40.713955826286046), "CA") %>%
   addHomeButton(extent(-88.33145141601564, -87.23281860351564, 41.466399253078876, 41.97276436226528), "IL") %>%
   addHomeButton(extent(-80.82092285156251, -79.72229003906251, 25.517657429994035, 26.12831612064242), "FL") %>%
   addHomeButton(extent(-73.95446777343751, -71.75720214843751, 41.03793062246529, 42.05337156043361), "CT") %>%
   addHomeButton(extent(-74.46807861328126, -73.91876220703126, 40.57484790030712, 40.83199550584334), "NJ2") %>%
   addHomeButton(extent(-75.86883544921876, -73.67156982421876, 39.16414104768742, 40.20824570152502), "NJ1") %>%
   addHomeButton(extent(-76.95098876953126, -74.75372314453126, 38.10214399750345, 39.16201148082406), "MD2") %>%
   addHomeButton(extent(-77.1947479248047, -76.6454315185547, 38.791021386961596, 39.0549177529185), "MD1") %>%
   addHomeButton(extent(-131.74804687500003, -61.43554687500001, 18.812717856407776, 52.908902047770255 ), "U.S.") %>%
   
   addLegend("bottomleft", 
             pal = binpal, 
             values = high_negative_equity_county_2019_geo$pct_negative_equity_y2019,
    title = "% Homes Negative Equity 2019",
    labFormat = labelFormat(prefix = ""),
    opacity = 1
  ) %>%
  addEasyButton(easyButton(
    states = list(
      easyButtonState(
        stateName="unfrozen-markers",
        icon="ion-toggle",
        title="Get Bounding box",
        onClick = JS("
                     function(btn, map) {
                        alert(
                         map.getBounds().getWest() + ', ' + map.getBounds().getEast() + ', ' + map.getBounds().getSouth() + ', ' + map.getBounds().getNorth() 
                        );
                        
                     }")
      )
    )
  )
)

```

```{r}

datatable(high_negative_equity_county_2019, class = 'cell-border stripe', filter = 'top', rownames = FALSE, caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    '2019: Counties with negative equity >= 4%'
  ),
  options = list(scrollX=TRUE, scrollCollapse=TRUE)
)
```

By comparison, this is every county in the U.S. in 2009 with a negative equity rate above 4 percent, 654 counties.  That's about 25 percent of counties.  The problem was much more widespread following the financial crisis. 

```{r}

# Filter for 2009
high_negative_equity_county_2009 <- underwater_county_year_no_zillow %>%
  #dplyr::select(state_code, fips_code, state_name, county_name, y2009) %>%
  filter(pct_negative_equity_y2009 >= 4) %>%
  arrange(desc(pct_negative_equity_y2009))

# Join zip code coordinates to negative equity by zipcode
high_negative_equity_county_2009_geo <- geo_join(counties, high_negative_equity_county_2009, 'GEOID', 'fips_code', 
how = "inner")  

# Color Scheme
binpal <- colorBin("plasma", high_negative_equity_county_2009_geo$pct_negative_equity_y2009, 5, pretty = FALSE)

# Draw map
leaflet(high_negative_equity_county_2009_geo) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%
   #addProviderTiles(providers$Wikimedia) %>%
   addPolygons(fillColor = ~binpal(pct_negative_equity_y2009), weight = 1, smoothFactor = 0.5, opacity = 0.1, fillOpacity = 0.5, color="black", popup = popupTable(high_negative_equity_county_2009_geo)) %>%
   setView(-95, 39.335359608681216, 4) %>%
   addLegend("bottomleft", 
             pal = binpal, 
             values = high_negative_equity_county_2009_geo$pct_negative_equity_y2009,
    title = "% Homes Negative Equity 2009",
    labFormat = labelFormat(prefix = ""),
    opacity = 1
  ) %>%
  addEasyButton(easyButton(
    states = list(
      easyButtonState(
        stateName="unfrozen-markers",
        icon="ion-toggle",
        title="Get Bounding box",
        onClick = JS("
                     function(btn, map) {
                        alert(
                         map.getBounds().getWest() + ', ' + map.getBounds().getEast() + ', ' + map.getBounds().getSouth() + ', ' + map.getBounds().getNorth() 
                        );
                        
                     }")
      )
    )
  )
)

```

```{r}

datatable(high_negative_equity_county_2009, class = 'cell-border stripe', filter = 'top', rownames = FALSE, caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    '2009: Counties with negative equity >= 4%'
  ),
  options = list(scrollX=TRUE, scrollCollapse=TRUE)
)
```

This map shows 2009 negative equity rates in counties to shows only the top 89 counties in 2009 (the same number of counties with > 4 percent negative equity in 2019). It allows us to see which places were the most problematic in 2009. 
```{r}
# Filter for 2009
top_negative_equity_county_2009 <- underwater_county_year_no_zillow %>%
  #dplyr::select(state_code, fips_code, state_name, county_name, y2009) %>%
  arrange(desc(pct_negative_equity_y2009)) %>%
  head(n=89)

# Join zip code coordinates to negative equity by zipcode
top_negative_equity_county_2009_geo <- geo_join(counties, top_negative_equity_county_2009, 'GEOID', 'fips_code', 
how = "inner")  

# Color Scheme
binpal <- colorBin("plasma", top_negative_equity_county_2009_geo$pct_negative_equity_y2009, 5, pretty = FALSE)

# Draw map
leaflet(top_negative_equity_county_2009_geo) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%
   #addProviderTiles(providers$Wikimedia) %>%
   addPolygons(fillColor = ~binpal(pct_negative_equity_y2009), weight = 1, smoothFactor = 0.5, opacity = 0.1, fillOpacity = 0.5, color="black", popup = popupTable(high_negative_equity_county_2009_geo)) %>%
   setView(-95, 39.335359608681216, 4) %>%
   addLegend("bottomleft", 
             pal = binpal, 
             values =top_negative_equity_county_2009_geo$pct_negative_equity_y2009,
    title = "% Homes Negative Equity 2009",
    labFormat = labelFormat(prefix = ""),
    opacity = 1
  ) %>%
  addEasyButton(easyButton(
    states = list(
      easyButtonState(
        stateName="unfrozen-markers",
        icon="ion-toggle",
        title="Get Bounding box",
        onClick = JS("
                     function(btn, map) {
                        alert(
                         map.getBounds().getWest() + ', ' + map.getBounds().getEast() + ', ' + map.getBounds().getSouth() + ', ' + map.getBounds().getNorth() 
                        );
                        
                     }")
      )
    )
  )
)

```
```{r}

datatable(top_negative_equity_county_2009, class = 'cell-border stripe', filter = 'top', rownames = FALSE, caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    '2009: 89 Counties with highest negative equity rates'
  ),
  options = list(scrollX=TRUE, scrollCollapse=TRUE)
)
```

There's not a lot of crossover. Only 13 counties on the list of the highest negative equity counties in 2009 were also there in 2019, including two Maryland counties.  

```{r}

on_both <- top_negative_equity_county_2009 %>%
  inner_join(high_negative_equity_county_2019, by=c("fips_code", "county_name", "state_name", "pct_negative_equity_y2009", "pct_negative_equity_y2019")) %>%
  dplyr::select(fips_code, county_name, state_name, pct_negative_equity_y2009, pct_negative_equity_y2019)


datatable(on_both, class = 'cell-border stripe', filter = 'top', rownames = FALSE, caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    'County with highest rates in 2019 and 2009'
  ),
  options = list(scrollX=TRUE,scrollY=TRUE, scrollCollapse=TRUE, pageLength = 13)
)
```


## Historical Comparisons

In the average and median U.S. county, underwater rates nowhere near as high as they used to be in the years immediately following the crash of the housing market. In 2009, the average U.S. county had 5.5 percent of homes with negative equity, compared to 1.52 percent today.  The median showed a less dramatic decline. The 2009 mean is skewed up by some extremely high rates in select counties in 2009.  

```{r}

x <- underwater_county_year_no_zillow %>%
  pivot_longer(cols=c("pct_negative_equity_y2009", "pct_negative_equity_y2010", "pct_negative_equity_y2011", "pct_negative_equity_y2012","pct_negative_equity_y2013","pct_negative_equity_y2014","pct_negative_equity_y2015","pct_negative_equity_y2016","pct_negative_equity_y2017","pct_negative_equity_y2018","pct_negative_equity_y2019"), names_to = "year", values_to = "percent_negative_equity") %>%
  group_by(year) %>%
  filter(!is.na(percent_negative_equity)) %>%
  summarise(mean_percent_negative_equity = round(mean(percent_negative_equity),2),
    median_percent_negative_equity = round(median(percent_negative_equity),2)
            )

datatable(x, class = 'cell-border stripe', filter = 'top', rownames = FALSE, caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    'Mean and median negative equity county rates by year 2009 and 2019'
  ),
  options = list(scrollX=TRUE,scrollY=TRUE, scrollCollapse=TRUE, pageLength = 11)
)
```


## Explaining the differences
In general, amongst the highest negative equity zips, there's a general trend.  The higher the rate of af american and hispanics, the higher the rate.

```{r}

# Average table
underwater_zips_2019_no_zillow %>%
  mutate(negative_equity_group = case_when(percent_negative_equity > 4 ~ "greater than 4",
                                      TRUE ~ "less than 4")
         ) %>%
  filter(!is.na(pct_black_2017), !is.na(median_household_income_2017)) %>%
  group_by(negative_equity_group) %>%
  summarise(pct_black_2017 = round(mean(pct_black_2017),2),
            pct_white_2017 = round(mean(pct_white_2017),2),
            pct_hispanic_2017 = round(mean(pct_hispanic_2017),2),
            median_household_income_2017 = round(mean(median_household_income_2017),2)) %>%
  datatable(class = 'cell-border stripe', filter = 'top', rownames = FALSE, caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left;',
    'Averages'
  ),
  options = list(scrollX=TRUE,scrollY=TRUE, scrollCollapse=TRUE, pageLength = 11)
)
```

And amongst higher negative equity ZIP codes (>= 4%), we see significant trends towards higher rates of negative equity as ZIP codes' in more African-American and Hispanic areas.   

```{r}
# Correlation for higher rate counties
underwater_zips_2019_no_zillow_trimmed <- underwater_zips_2019_no_zillow %>%
  dplyr::select(-zip_code,-state_name,-county_name) %>%
  filter(percent_negative_equity >=4)  

# Create correlation matrix
underwater_zips_2019_no_zillow_correlation_matrix <- rcorr(as.matrix(underwater_zips_2019_no_zillow_trimmed))

# Flatten correlation table, to only include significant predictors p < .05
underwater_zips_2019_no_zillow_correlation_matrix <- flattenCorrMatrix(underwater_zips_2019_no_zillow_correlation_matrix$r, underwater_zips_2019_no_zillow_correlation_matrix$P) %>%
  mutate(absolute_cor = abs(cor)) %>%
  filter(row == "percent_negative_equity") %>%
  filter(p < .05) %>%
  arrange(desc(absolute_cor)) %>%
  dplyr::select(-absolute_cor)

# Print correlation matrix.  Let's us see both r (correlation coefficient) and p value for each predictor on an individual level
underwater_zips_2019_no_zillow_correlation_matrix %>%
  filter(str_detect(column,"white|black|hispanic")) %>%
  datatable()

```
There is correlation coefficient of .33 between an areas' percentage of African-Americans' and its negative equity rate, a moderate positive relationship. This is not causal, per se, just that a third of the variance in negative equity rates is explained by an areas' concentration of African Americans. Areas with higher rates of African-Americans tend to have higher negative equity rates.  
```{r}


ggscatter(underwater_zips_2019_no_zillow_trimmed, x="percent_negative_equity",y= "pct_black_2017", add="loess", add.params = list(color = "blue", fill = "lightgray"), conf.int=TRUE) +
  stat_cor(method = "pearson", label.x =15, label.y = 15)
```
There is correlation coefficient of .30 between an areas' percentage of Hispanics and its negative equity rate, a moderate positive relationship. This is not causal, per se, just that a third of the variance in negative equity rates is explained by an areas' concentration of African Americans.  Areas with higher rates of Hispanics tend to have higher negative equity rates. 
```{r}


ggscatter(underwater_zips_2019_no_zillow_trimmed, x="percent_negative_equity",y= "pct_hispanic_2017", add="loess", add.params = list(color = "blue", fill = "lightgray"), conf.int=TRUE) +
  stat_cor(method = "pearson", label.x = 16, label.y = 15)

```

As one might expect, the opposite is true for whiter areas. There is correlation coefficient of -.4 between an areas' percentage of whites and its negative equity rate, a moderate negative relationship. This is not causal, per se, just that a two-fifths of the variance in negative equity rates is explained by an areas' concentration of whites.  Areas with higher rates of whites tend to have higher lower negative equity rates. 

```{r}


ggscatter(underwater_zips_2019_no_zillow_trimmed, x="percent_negative_equity",y= "pct_white_2017", add="loess", add.params = list(color = "blue", fill = "lightgray"), conf.int=TRUE) +
  stat_cor(method = "pearson", label.x = 16, label.y = 15)

```

## Examples





































































f
## Load and clean zillow data from data.world
Strategy
What are the ramifications of negative equity?

Negative equity can cast a shadow over the economy, effectively prolonging an economic downturn. Just as skyrocketing home values can drive confidence and spending, negative equity can keep people home, literally. Many people who might have found jobs in other markets were tied to their homes and unable to sell because the money they would make from selling would not be enough to pay off their mortgages. Negative equity can have a number of other chilling impacts on local housing markets, disproportionately impacting minority communities and owners of lower-valued homes, exacerbating inventory shortages and increasing the likelihood of foreclosure for underwater homeowners.
# https://www.zillow.com/research/negative-equity-race-q3-2016-14063/
# https://www.zillow.com/research/q3-2016-negative-equity-report-13954/


https://www.zillow.com/research/q3-2016-negative-equity-report-13954/
But this drop, while encouraging, masks the often very wide divide that remains between the top of the market and the bottom of the market. In a number of large markets, the spread between the negative equity rate at the top and the bottom of the market is alarmingly wide. In Detroit, for example, the negative equity rate among top-tier homes is 4.4 percent; among bottom-tier homes, the negative equity rate is almost ten times higher at 39 percent – a gap of 34.6 percentage points, the largest gap among the 35 largest metros analyzed. The gap between top-tier and bottom-tier negative equity is more than 20 percentage points in an additional four large metros: Cleveland (28.1 points), St. Louis (22 points), Atlanta (21 points) and Chicago (20.5 points).



```{r}
# https://datadotworld.github.io/data.world-r/r-rstudio.html
# install.packages('data.world')
library(data.world)

# Store token
saved_cfg <- data.world::save_config("eyJhbGciOiJIUzUxMiJ9.eyJzdWIiOiJyLWFuZC1yLXN0dWRpbzpzbXVzc2VuZGVuIiwiaXNzIjoiY2xpZW50OnItYW5kLXItc3R1ZGlvOmFnZW50OnNtdXNzZW5kZW46OmE1YWFmYjRhLTM4NTItNDc4Mi05Y2JjLTgzMDVmN2UyZWE5MCIsImlhdCI6MTU3NDQ1NzczMCwicm9sZSI6WyJ1c2VyX2FwaV9hZG1pbiIsInVzZXJfYXBpX3JlYWQiLCJ1c2VyX2FwaV93cml0ZSJdLCJnZW5lcmFsLXB1cnBvc2UiOnRydWUsInNhbWwiOnt9fQ.9dc8h4PyGmm3CQcEgoG8fTFsA-Qz8nYNkCd39CiWRrUQV5slhWNBGra82yFQhQuwCfvdfZlL06utF1zaJxoNlw")
data.world::set_config(saved_cfg)
# https://data.world/zillow-data/all-zillow-metrics-zip-code/workspace/file?filename=Zip%2FBuyerSellerIndex_Zip.csv
# Define path to dataset
negative_equity_summary_url <- "https://data.world/zillow-data/negative-equity-summary"
# Define query to load dataset
negative_equity_summary_query <- data.world::qry_sql("SELECT * FROM NESummary_2017Q1_Public")
# Get data
negative_equity_summary <- data.world::query(negative_equity_summary_query, dataset = negative_equity_summary_url)
# Filter by county
negative_equity_summary_county <- negative_equity_summary %>%
  filter(regiontype == "County")
# Filter by zip
negative_equity_summary_zip <- negative_equity_summary %>%
  filter(regiontype == "Zip")

# write.csv(negative_equity_summary_zip, "negative_equity_summary_zip.csv")

#### NEXT DATA SET

market_health_zip_url <- "https://data.world/zillow-data/all-zillow-metrics-zip-code/"
# Define query to load dataset
market_health_zip_query <- data.world::qry_sql(" SELECT * FROM markethealthindex_zip")
# Get data
market_health_zip_summary <- data.world::query(market_health_zip_query, dataset = market_health_zip_url)





https://data.world/zillow-data/all-zillow-metrics-zip-code/workspace/file?filename=Zip%2FBuyerSellerIndex_Zip.csv





# https://data.world/zillow-data/negative-equity-tiers
#https://data.world/zillow-data/negative-equity-summary
#https://data.world/zillow-data/negative-equity-time-series
#https://data.world/zillow-data/negative-equity-summary
#HOLY SHIT THERE S A FUCKING MODEL ON PAGE 14 that has everything I need. #https://www.frbatlanta.org/-/media/documents/community-development/publications/discussion-papers/2016/01-housing-negative-equity-in-sixth-federal-reserve-district-2016-03-10.pdf

# ZIP Codes

# Get list of states
states <- state.abb 

# Take out texas, Vermont
states <- states[!states %in% c("TX", "VT", "WY")]  
  
# Just for testing
# states <- c("AL", "AK")
# state <- "CT"

negative_equity_summary_acs_zip <- negative_equity_summary_zip %>%
  inner_join(acs_data_zcta, by=c("regionname" = "geoid")) %>%
  mutate(majority_minority = case_when(pct_white < 50 ~ TRUE,
                                       TRUE ~ FALSE))


# Make an empty dataframe
negative_equity_acs_zip_regressions <- tibble(
  state_eff_neg_eq= character(), 
  pct_white = double(),
  pct_black = double(),
  pct_hispanic = double(),
  median_household_income = double(),
  pct_poverty = double()
  )

# Make an empty dataframe 

negative_equity_acs_zip_averages <- tibble(
 state =character(),
 majority_minority =logical(),
 regionid=double(),
 sizerank=double(),
 total_amount_of_negative_equity_millions=double(),
 total_number_of_homes_in_negative_equity=double(), 
 negative_equity_percent_of_homes_w_mortgage =double(), 
 number_of_homes_at_least_90_days_late =double(), 
 negative_equity_deliquency_rate =double(), 
 `2017q1_zhvi` =double(), 
 zhvi_change_from_peak =double(), 
 ne_ltv_of_100_120 =double(), 
 ne_ltv_of_120_140 =double(), 
 ne_ltv_of_140_160 =double(), 
 ne_ltv_of_160_180 =double(), 
 ne_ltv_of_180_200 =double(), 
 ne_ltv_of_200 =double(), 
 ltv_of_40 =double(), 
 ltv_of_40_60=double(), 
 ltv_of_60_80=double(),
 ltv_of_80_100 =double(), 
 ltv_of_100_120=double(), 
 ltv_of_120_140=double(),
 ltv_of_140_160=double(),
 ltv_of_160_180=double(),
 ltv_of_180_200=double(),
 ltv_of_200=double(), 
 effective_negative_equity_rate=double(), 
 pct_white =double(),
 pct_black =double(), 
 pct_hispanic=double(),
 pct_poverty =double(),
 median_household_income =double()
)

# Make empty dataframe
negative_equity_acs_zip_models <- tibble(
state=character(),
r.squared=double(),
adj.r.squared=double(),
p.value=character(),
intercept=character(),
pct_white=character(),
pct_black=character(),
pct_poverty=character(),
median_household_income=character(),
pct_hispanic=character()
)

# Run the for loop
for (state in states) {

  print(state)
  
  temp <- negative_equity_summary_acs_zip %>%
  filter(state==!!enquo(state)) %>%
  select_if(is.numeric) %>%
  correlate() %>%
  dplyr::select(rowname, pct_white, pct_black, pct_hispanic, median_household_income, pct_poverty) %>%
  mutate(abs_pct_white=abs(pct_white)) %>%
  arrange(desc(abs_pct_white)) %>%
  dplyr::select(-abs_pct_white) %>%
  filter(rowname == "effective_negative_equity_rate") %>%
  mutate(state_eff_neg_eq = state) %>%
  dplyr::select(state_eff_neg_eq, everything(), -rowname)
  
  negative_equity_acs_zip_regressions <- negative_equity_acs_zip_regressions %>%
      bind_rows(temp)
  
  temp2 <- negative_equity_summary_acs_zip  %>%
  filter(state==!!enquo(state)) %>%  
  group_by(majority_minority) %>%
  summarise_if(is.numeric, mean, na.rm=TRUE) %>%
  mutate(state = state) %>%
  dplyr::select(state, everything())
  
  negative_equity_acs_zip_averages <-  negative_equity_acs_zip_averages %>%
    bind_rows(temp2) 
  

  # Filter out each state
temp3 <- negative_equity_summary_acs_zip %>%
  filter(state==!!enquo(state)) 

# Build a linear model with all of our variables
temp3 <- lm(effective_negative_equity_rate~pct_white+pct_black+pct_poverty+median_household_income+pct_hispanic, data=temp3)

# Create a tidy dataframe with significance of each variable in a LM
temp3_tidy <- tidy(temp3) %>%
  mutate(state=!!enquo(state)) %>%
  dplyr::select(state, everything(), -estimate, -std.error,-statistic) %>%
  mutate(p.value = case_when(p.value < .05 ~ "sig",
                             TRUE ~ 'not sig')) %>%
  pivot_wider(names_from = term, values_from = p.value) %>%
  rename(intercept = `(Intercept)`)


# Create a tidy dataframe with overall performance of model
temp3_glance <- glance(temp3) %>%
  mutate(state=!!enquo(state)) %>%
  dplyr::select(state, r.squared, adj.r.squared, p.value) %>%
  mutate(p.value = case_when(p.value < .05 ~ "sig",
                             TRUE ~ 'not sig'))

# Join the two together
temp3_join <- temp3_glance %>%
  inner_join(temp3_tidy, by=c("state"))

# Bind to our empty dataframe
negative_equity_acs_zip_models <- negative_equity_acs_zip_models %>%
  bind_rows(temp3_join)

}

# Filter out each state
temp3 <- negative_equity_summary_acs_zip %>%
  filter(state==!!enquo(state)) 

# Build a linear model with all of our variables
temp3 <- lm(effective_negative_equity_rate~pct_white+pct_black+pct_poverty+median_household_income+pct_hispanic, data=temp3)
  
# Create a tidy dataframe with significance of each variable in a LM
temp3_tidy <- tidy(test) %>%
  mutate(state=!!enquo(state)) %>%
  dplyr::select(state, everything(), -estimate, -std.error,-statistic) %>%
  mutate(p.value = case_when(p.value < .05 ~ "sig",
                             TRUE ~ 'not sig')) %>%
  pivot_wider(names_from = term, values_from = p.value) %>%
  rename(intercept = `(Intercept)`)


# Create a tidy dataframe with overall performance of model
temp3_glance <- glance(test) %>%
  mutate(state=!!enquo(state)) %>%
  dplyr::select(state, r.squared, adj.r.squared, p.value) %>%
  mutate(p.value = case_when(p.value < .05 ~ "sig",
                             TRUE ~ 'not sig'))

# Join the two together
temp3_join <- temp3_glance %>%
  inner_join(temp3_tidy, b=c("state"))

# Bind to our empty dataframe
negative_equity_acs_zip_models <- negative_equity_acs_zip_models %>%
  bind_rows(temp3_join)



x <- negative_equity_summary_zip %>%
  inner_join(acs_data_zcta, by=c("regionname" = "geoid")) %>%
  mutate(majority_minority = case_when(pct_white < 50 ~ TRUE,
                                       TRUE ~ FALSE)) #%>%
  #filter(state=="MD")
y <- x %>%
  group_by(majority_minority) %>%
  summarise_if(is.numeric, mean, na.rm=TRUE)

# Correlate

x %>%
  select_if(is.numeric) %>%
  correlate() %>%
  dplyr::select(rowname, pct_white, pct_black, pct_hispanic, median_household_income, pct_poverty) %>%
  mutate(abs_pct_white=abs(pct_white)) %>%
  arrange(desc(abs_pct_white)) %>%
  dplyr::select(-abs_pct_white) %>%
  filter(rowname == "effective_negative_equity_rate")
  

negative_equity_summary_zip 
# Percent black
# Percent non-white hispanic
# Median Household Income
# Mean commute time
# Owner occupied rate
#Averate age of housing - don't have
#Homeowner vacancy rate - don't have
#rental vacancy rate  - don't have
# Percent subprime loans at peak - don't have
# Percent foreclosure rate - don't have
# Home price decline from peak - HAVE ZHVI change from peak


# Join to county data  

x <- negative_equity_summary_county %>%
  inner_join(zillow_crosswalk, by=c("regionid" = "CountyRegionID_Zillow")) %>%
  dplyr::select(fips_code, everything()) %>%
  inner_join(acs_data_county, by=c("fips_code" = "geoid")) %>%
  mutate(majority_minority = case_when(pct_white < 75 ~ TRUE,
                                       TRUE ~ FALSE))

# Averages

y <- x %>%
  group_by(majority_minority) %>%
  summarise_if(is.numeric, mean, na.rm=TRUE)

# Correlate

x %>%
  select_if(is.numeric) %>%
  correlate() %>%
  dplyr::select(rowname, pct_white, pct_black, pct_hispanic, median_household_income) %>%
  mutate(abs_pct_white=abs(pct_white)) %>%
  arrange(desc(abs_pct_white)) %>%
  dplyr::select(-abs_pct_white)
select_i

```

## Load Shapefiles

```{r}
# ZIP Code Points
data(zipcode)

options(tigris_use_cache=TRUE)

# ZCTA shapefiles
zctas <- zctas(cb=TRUE)

# Counties
counties <- counties(cb = TRUE)

```

## Explaining the differences
In general, amongst the highest negative equity zips, there's a general trend.  The higher the rate of af american and hispanics, the higher the rate.

```{r}

# Remove non-numerical columns, and eliminate less significant negative equity zips
zcta_uw_census_trimmed <- zcta_uw_census %>%
  dplyr::select(-zip_code,-state_name,-county_name) %>%
  filter(percent_negative_equity >=5) %>%
  mutate(log_median_household_income = log(median_household_income))

# Create correlation matrix
zcta_uw_census_correlation_matrix <- rcorr(as.matrix(zcta_uw_census_trimmed))

# Flatten correlation table, to only include significant predictors p < .05
zcta_uw_census_correlation_matrix <- flattenCorrMatrix(zcta_uw_census_correlation_matrix$r, zcta_uw_census_correlation_matrix$P) %>%
  mutate(absolute_cor = abs(cor)) %>%
  filter(row == "percent_negative_equity") %>%
  filter(p < .05) %>%
  arrange(desc(absolute_cor)) %>%
  dplyr::select(-absolute_cor)

# Print correlation matrix.  Let's us see both r (correlation coefficient) and p value for each predictor on an individual level
zcta_uw_census_correlation_matrix %>%
  datatable()

# Create multiple linear model to explain (creating two, one with all variables and one with selected variables)

# All variables
zcta_uw_census_linear_model_all <- lm(percent_negative_equity ~ ., data = zcta_uw_census_trimmed)

# Print a summary of the model with all variables
summary(zcta_uw_census_linear_model_all)

# Print a regression table and summary of all variables
get_regression_table(zcta_uw_census_linear_model_all) %>% 
  datatable()
get_regression_summaries(zcta_uw_census_linear_model_all) %>%
  mutate(r = sqrt(r_squared)) %>% 
  dplyr::select(r_squared, adj_r_squared, r, everything()) %>%
  datatable() 

# Select variables
zcta_uw_census_linear_model_select <- lm(percent_negative_equity ~ pct_white+pct_hispanic+five_year_pct_change+ten_year_pct_change, data = zcta_uw_census_trimmed)

# Print a summary of the model with select variables
summary(zcta_uw_census_linear_model_select)

# Print a regression table of select variables
get_regression_table(zcta_uw_census_linear_model_select) %>% 
  datatable()
get_regression_summaries(zcta_uw_census_linear_model_select) %>%
  mutate(r = sqrt(r_squared)) %>% 
  dplyr::select(r_squared, adj_r_squared, r, everything()) %>%
  datatable() 

# Plot one property using ggpubr scatter package

# Just as a reminder, here's what a perfect correlation plot looks like.  Tightly grouped around the prediction line of best fit.
ggscatter(zcta_uw_census_trimmed, x="percent_negative_equity", "percent_negative_equity", add="reg.line", add.params = list(color = "blue", fill = "lightgray"), conf.int=TRUE) +
  stat_cor(method = "pearson", label.x = 3, label.y = 15)

# Make facet graphs
graph_data <- zcta_uw_census_trimmed %>%
  gather(-percent_negative_equity, key = "var", value = "value") %>%
  inner_join(zcta_uw_census_correlation_matrix, by=c("var" = "column")) %>%
  dplyr::select(-row, -cor, -p) %>%
  ggscatter(x="value", y="percent_negative_equity", add="loess", add.params = list(color = "blue", fill = "lightgray"), conf.int=TRUE, mean.point = TRUE) +
  stat_cor(method = "pearson", label.x = 0, label.y = 25)

facet(graph_data, facet.by="var", scales="free_x", ncol=4)

```

```{r}

# Data.world in R -- lets build a model for q12017 that explains everything.  Use it to update for now. 


# https://data.world/zillow-data
# Let's build a model using this rate https://data.world/zillow-data/negative-equity-summary
# https://data.world/zillow-data/negative-equity-tiers
# https://www.census.gov/content/dam/Census/programs-surveys/ahs/publications/Drowning_in_Debt.pdf
# Owner to renter
# https://data.world/zillow-data/negative-equity-time-series

# Remove non-numerical columns, and eliminate less significant negative equity zips
county_uw_ts_census_rural_trimmed <- county_uw_ts_census_rural %>%
  dplyr::select(-fips_code,-state_name,-county_name, -pct_negative_equity_y2010, -pct_negative_equity_y2011, -pct_negative_equity_y2012, -pct_negative_equity_y2013, -pct_negative_equity_y2014, -pct_negative_equity_y2015, -pct_negative_equity_y2016, -pct_negative_equity_y2017, -pct_negative_equity_y2018) 
  # %>% filter(pct_negative_equity_y2019 >=5) 

# Create correlation matrix
county_uw_ts_census_rural_correlation_matrix <- rcorr(as.matrix(county_uw_ts_census_rural_trimmed))

# Flatten correlation table, to only include significant predictors p < .05
county_uw_ts_census_rural_correlation_matrix <- flattenCorrMatrix(county_uw_ts_census_rural_correlation_matrix$r, county_uw_ts_census_rural_correlation_matrix$P) %>%
  mutate(absolute_cor = abs(cor)) %>%
  filter(row == "pct_negative_equity_y2019") %>%
  filter(p < .05) %>%
  arrange(desc(absolute_cor)) %>%
  dplyr::select(-absolute_cor)

# Print correlation matrix.  Let's us see both r (correlation coefficient) and p value for each predictor on an individual level
county_uw_ts_census_rural_correlation_matrix %>%
  datatable()

# Create multiple linear model to explain (creating two, one with all variables and one with selected variables)

# All variables
county_uw_ts_census_rural_linear_model_all <- lm(pct_negative_equity_y2019 ~ ., data = county_uw_ts_census_rural_trimmed)

# Print a summary of the model with all variables
summary(county_uw_ts_census_rural_linear_model_all)

# Print a regression table and summary of all variables
get_regression_table(county_uw_ts_census_rural_linear_model_all) %>% 
  datatable()

get_regression_summaries(county_uw_ts_census_rural_linear_model_all) %>%
  mutate(r = sqrt(r_squared)) %>% 
  dplyr::select(r_squared, adj_r_squared, r, everything()) %>%
  datatable() 

# Select variables
county_uw_ts_census_rural_linear_model_select <- lm(pct_negative_equity_y2019 ~ pct_white+pct_negative_equity_y2009+pct_poverty+pct_black+pct_hispanic, data = county_uw_ts_census_rural_trimmed)

# Print a summary of the model with select variables
summary(county_uw_ts_census_rural_linear_model_select)

# Print a regression table of select variables
get_regression_table(county_uw_ts_census_rural_linear_model_select) %>% 
  datatable()

get_regression_summaries(county_uw_ts_census_rural_linear_model_select) %>%
  mutate(r = sqrt(r_squared)) %>% 
  dplyr::select(r_squared, adj_r_squared, r, everything()) %>%
  datatable() 

# Plot one property using ggpubr scatter package

# Just as a reminder, here's what a perfect correlation plot looks like.  Tightly grouped around the prediction line of best fit.
ggscatter(zcta_uw_census_trimmed, x="percent_negative_equity", "percent_negative_equity", add="reg.line", add.params = list(color = "blue", fill = "lightgray"), conf.int=TRUE) +
  stat_cor(method = "pearson", label.x = 3, label.y = 15)

# Make facet graphs
graph_data <- zcta_uw_census_trimmed %>%
  gather(-percent_negative_equity, key = "var", value = "value") %>%
  inner_join(zcta_uw_census_correlation_matrix, by=c("var" = "column")) %>%
  dplyr::select(-row, -cor, -p) %>%
  ggscatter(x="value", y="percent_negative_equity", add="reg.line", add.params = list(color = "blue", fill = "lightgray"), conf.int=TRUE, mean.point = TRUE) +
  stat_cor(method = "pearson", label.x = 0, label.y = 25)

facet(graph_data, facet.by="var", scales="free_x", ncol=3)
```

```{r}

```



```{r}
county_uw_ts_census_rural %>%
  dplyr::select(-fips_code,-state_name,-county_name) %>%
  #filter(pct_negative_equity_y2019 >=1) %>%
  correlate() %>%
  dplyr::select(rowname, pct_negative_equity_y2019) %>%
  arrange(desc(pct_negative_equity_y2019))




library("Hmisc")

x <- county_uw_ts_census_rural %>%
  dplyr::select(-fips_code,-state_name,-county_name) %>%
  filter(pct_negative_equity_y2019 >=1) 

#res2 <- rcorr(as.matrix(x))
#res2
#y <- as_tibble(res2$r)
#y
#res2$P
res2 <- rcorr(as.matrix(x))

flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}

y <- flattenCorrMatrix(res2$r, res2$P) %>%
  mutate(absolute_cor = abs(cor)) %>%
  filter(row == "percent_negative_equity") %>%
  filter(p < .05) %>%
  arrange(desc(absolute_cor)) %>%
  dplyr::select(-absolute_cor)

y

score_model <- lm(pct_negative_equity_y2019 ~ ., data = x)
summary(score_model)

#score_model

get_regression_table(score_model) %>% datatable()
get_regression_summaries(score_model) %>% datatable()


```
R-squared is a statistical measure of how close the data are to the fitted regression line. It is also known as the coefficient of determination, or the coefficient of multiple determination for multiple regression.

The definition of R-squared is fairly straight-forward; it is the percentage of the response variable variation that is explained by a linear model. Or:

R-squared = Explained variation / Total variation

R-squared is always between 0 and 100%:

    0% indicates that the model explains none of the variability of the response data around its mean.
    100% indicates that the model explains all the variability of the response data around its mean.

In general, the higher the R-squared, the better the model fits your data. However, there are important conditions for this guideline that I’ll talk about both in this post and my next post.
```{r}
county_uw_2019_census_rural_zillow %>%
  dplyr::select(-fips_code,-state_name,-county_name) %>%
  #filter(pct_negative_equity_y2019 >=1) %>%
  correlate() %>%
  dplyr::select(rowname, pct_negative_equity_y2019) %>%
  arrange(desc(pct_negative_equity_y2019))



library("Hmisc")

x <- county_uw_2019_census_rural_zillow %>%
  dplyr::select(-fips_code,-state_name,-county_name) #%>%
 # filter(pct_negative_equity_y2019 >=1) 

#res2 <- rcorr(as.matrix(x))
#res2
#y <- as_tibble(res2$r)
#y
#res2$P
res2 <- rcorr(as.matrix(x))

flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}

y <- flattenCorrMatrix(res2$r, res2$P) %>%
  mutate(absolute_cor = abs(cor)) %>%
  filter(row == "percent_negative_equity") %>%
  filter(p < .05) %>%
  arrange(desc(absolute_cor)) %>%
  dplyr::select(-absolute_cor)

y

score_model <- lm(pct_negative_equity_y2019 ~ ., data = x)
summary(score_model)

#score_model

get_regression_table(score_model) %>% datatable()
get_regression_summaries(score_model) %>% datatable()

```

```{r}
county_uw_2019_census_rural_zillow_dq %>%
  dplyr::select(-fips_code,-state_name,-county_name) %>%
  #filter(pct_negative_equity_y2019 >=1) %>%
  correlate() %>%
  dplyr::select(rowname, pct_negative_equity_y2019) %>%
  arrange(desc(pct_negative_equity_y2019))

library("Hmisc")

x <- county_uw_2019_census_rural_zillow_dq %>%
  dplyr::select(-fips_code,-state_name,-county_name) #%>%
 # filter(pct_negative_equity_y2019 >=1) 

#res2 <- rcorr(as.matrix(x))
#res2
#y <- as_tibble(res2$r)
#y
#res2$P
res2 <- rcorr(as.matrix(x))

flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}

y <- flattenCorrMatrix(res2$r, res2$P) %>%
  mutate(absolute_cor = abs(cor)) %>%
  filter(row == "percent_negative_equity") %>%
  filter(p < .05) %>%
  arrange(desc(absolute_cor)) %>%
  dplyr::select(-absolute_cor)

y

score_model <- lm(pct_negative_equity_y2019 ~ ., data = x)
summary(score_model)

#score_model

get_regression_table(score_model) %>% datatable()
get_regression_summaries(score_model) %>% datatable()


```

```{r}

### Need to add in home price appreciation and mortgage origination as percentage of application
# Pull in NYT classification of urban rural

underwater_zips_2019_join <- underwater_zips_2019 %>%
  inner_join(acs_data_zcta, by=c("zip_code" = "geoid")) %>%
  inner_join(Zip_ZHVI_summary_current_month, by=c("zip_code" = "RegionName")) %>% 
  #%>%
  #filter(!is.na(percent_negative_equity),
   #      pct_white != "NaN",
   #      !is.na(median_household_income)) %>%
  filter(percent_negative_equity >=5) 
 
corr_table_zcta <- underwater_zips_2019_join %>%
  dplyr::select(-matches("code|name|Date|Region|State|Metro|County|City|Month|Quarter|Last|Time")) %>%
  dplyr::select(matches("percent|pct|year")) %>%
  correlate() %>%
  dplyr::select(rowname, percent_negative_equity)

print(corr_table_zcta)
#https://www.dataquest.io/blog/statistical-learning-for-predictive-modeling-r/
#install.packages("moderndive")

score_model <- lm(percent_negative_equity ~ pct_white+pct_black+pct_hispanic+PctFallFromPeak, data = underwater_zips_2019_join)
get_regression_table(score_model) %>% datatable()
get_regression_summaries(score_model) %>% datatable()

ggplot(underwater_zips_2019_join) +
  geom_point(aes(percent_negative_equity, pct_white)) +
  labs(x="Percent Negative Equity", y="Percent White", title="", caption = "") +
  geom_smooth(aes(percent_negative_equity, pct_white), method = "lm", se = FALSE) 

ggplot(underwater_zips_2019_join) +
  geom_point(aes(percent_negative_equity, pct_black)) +
  labs(x="Percent Negative Equity", y="Percent Black", title="", caption = "") +
  geom_smooth(aes(percent_negative_equity, pct_black), method = "lm", se = FALSE) 

ggplot(underwater_zips_2019_join) +
  geom_point(aes(percent_negative_equity, pct_hispanic)) +
  labs(x="Percent Negative Equity", y="Percent Hispanic", title="", caption = "") +
  geom_smooth(aes(percent_negative_equity, pct_hispanic), method = "lm", se = FALSE) 



```

# Analysis

In the average and median U.S. county, underwater rates nowhere near as high as they used to be in the years immediately following the crash of the housing market. In 2009, the average U.S. county had 5.5 percent of homes with negative equity, compared to 1.52 percent today.  The median showed a less dramatic decline. The 2009 mean is skewed up by some extremely high rates in select counties in 2009.  

```{r}
underwater_county_year %>%
  pivot_longer(cols=c("y2009", "y2010", "y2011", "y2012","y2013","y2014","y2015","y2016","y2017","y2018","y2019"), names_to = "year", values_to = "percent_negative_equity") %>%
  group_by(year) %>%
  filter(!is.na(percent_negative_equity)) %>%
  summarise(median_percent_negative_equity = round(median(percent_negative_equity),2),
            mean_percent_negative_equity = round(mean(percent_negative_equity),2)
            )
```

```{r}
underwater_county_year %>%
  filter(y2009 > 0) %>%
  pivot_longer(cols=c("y2009", "y2010", "y2011", "y2012","y2013","y2014","y2015","y2016","y2017","y2018","y2019"), names_to = "year", values_to = "percent_negative_equity") %>%
  group_by(year) %>%
  filter(!is.na(percent_negative_equity)) %>%
  summarise(median_percent_negative_equity = round(median(percent_negative_equity),2),
            mean_percent_negative_equity = round(mean(percent_negative_equity),2)
            )
```

The distribution of counties shows this.  In 2009, there were lots of counties with negative equity rates above 20 percent.  In 2019, only a handful have rates higher than 10.

```{r}

underwater_county_year %>%
  pivot_longer(cols=c("y2009", "y2010", "y2011", "y2012","y2013","y2014","y2015","y2016","y2017","y2018","y2019"), names_to = "year", values_to = "percent_negative_equity") %>%
  filter(str_detect(year,"y2009|y2012|y2015|y2019")) %>%
  filter(percent_negative_equity > 0) %>%
  ggplot(aes(percent_negative_equity)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(~year)


#underwater_county_year %>%
#  pivot_longer(cols=c("y2009", "y2010", "y2011", "y2012","y2013","y2014","y2015","y2016","y2017","y2018","y2019"), names_to = "year", values_to = #"percent_negative_equity") %>%
#  group_by(year) %>%
#  filter(!is.na(percent_negative_equity)) %>%
#  summarise(median_percent_negative_equity = round(median(percent_negative_equity),2),
 #           mean_percent_negative_equity = round(mean(percent_negative_equity),2)
 #           )
```

## Negative Equity Clusters in 2019 

Though negative equity rates are nowhere near historical levels, the analysis identified 10 clusters with a higher negative equity rates. 

In the map below, click on the two-letter buttons at right to zoom to that cluster. 

Click on each ZIP code to see info about each county.

Clusters
* MD1 | Maryland+DC | Cluster of ZIP Codes in Prince George's County and Anacostia in D.C. and down through Waldorm, in Charles County.
* MD2 | Maryland Eastern Shore | Huge chunks of the lower Eastern Shore, through Salisbury and Ocean City.
* NJ1 | South New Jersey | Large swaths of southern New Jersey, including Atlantic City, Philly suburbs.
* NJ2 | North Jersey+NYC | New York City suburbs, including Newark, Elizabeth, Paterson, parts of Queens
* CT | Connecticut | Several areas, including Hartford and Waterbury, Bridgeport and New Haven, up to Rhode Island border.
* IA | Iowa | a bunch of communities in Iowa with no real clusters. Can't make heads or tails of this.
* FL | Florida | Worst issues in Miami, Hialeah and Homestead, scatterd parts of the state.
* IL | Chicago | Huge issues surrounding Chicago, especially on South Side
* CA | California | Issues in Fresno, south of Monterey, scattered throughout. 
* NV | Las Vegas | 
* No buttons, but interesting | Baton Rouge, New Orleans, Atlanta, North Dakota


```{r}

# Filter for only high negative equity zip codes

underwater_zips_2019_x <- underwater_zips_2019 %>%
  filter(percent_negative_equity >= 5)

# Join zip code coordinates to negative equity by zipcode
underwater_zctas_2019 <- geo_join(zctas, underwater_zips_2019_x, 'GEOID10', 'zip_code', 
how = "inner")

# Color Scheme
binpal <- colorBin("plasma", underwater_zctas_2019$percent_negative_equity, 5, pretty = FALSE)

# Draw map
leaflet(underwater_zctas_2019) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%
   #addProviderTiles(providers$Wikimedia) %>%
   addPolygons(fillColor = ~binpal(percent_negative_equity), weight = 1, smoothFactor = 0.5, opacity = 0.1, fillOpacity = 0.5, color="black", popup = popupTable(underwater_zctas_2019)) %>%
   setView(-95, 39.335359608681216, 4) %>%
   addHomeButton(extent(-98.39355468750001, -89.60449218750001, 39.85072092501597, 43.8899753738369), "IA") %>%
   addHomeButton(extent(-115.44021606445314, -114.89089965820314, 36.04521273039952, 36.318998009207924), "NV") %>%
   addHomeButton(extent(-127.24365234375001, -109.66552734375001, 31.98944183792288, 40.713955826286046), "CA") %>%
   addHomeButton(extent(-88.33145141601564, -87.23281860351564, 41.466399253078876, 41.97276436226528), "IL") %>%
   addHomeButton(extent(-80.82092285156251, -79.72229003906251, 25.517657429994035, 26.12831612064242), "FL") %>%
   addHomeButton(extent(-73.95446777343751, -71.75720214843751, 41.03793062246529, 42.05337156043361), "CT") %>%
   addHomeButton(extent(-74.46807861328126, -73.91876220703126, 40.57484790030712, 40.83199550584334), "NJ2") %>%
   addHomeButton(extent(-75.86883544921876, -73.67156982421876, 39.16414104768742, 40.20824570152502), "NJ1") %>%
   addHomeButton(extent(-76.95098876953126, -74.75372314453126, 38.10214399750345, 39.16201148082406), "MD2") %>%
   addHomeButton(extent(-77.1947479248047, -76.6454315185547, 38.791021386961596, 39.0549177529185), "MD1") %>%
   addHomeButton(extent(-131.74804687500003, -61.43554687500001, 18.812717856407776, 52.908902047770255 ), "U.S.") %>%
   addLogo("https://jeroenooms.github.io/images/banana.gif",
        position = "bottomleft",
        offset.x = 5,
        offset.y = 40,
        width = 100,
        height = 100) %>%
   addLegend("bottomleft", 
             pal = colorBin("plasma", underwater_zctas_2019$percent_negative_equity), 
             values = underwater_zctas_2019$percent_negative_equity,
    title = "% Homes Negative Equity 2019",
    labFormat = labelFormat(prefix = ""),
    opacity = 1
  ) %>%
  addEasyButton(easyButton(
    states = list(
      easyButtonState(
        stateName="unfrozen-markers",
        icon="ion-toggle",
        title="Get Bounding box",
        onClick = JS("
                     function(btn, map) {
                        alert(
                         map.getBounds().getWest() + ', ' + map.getBounds().getEast() + ', ' + map.getBounds().getSouth() + ', ' + map.getBounds().getNorth() 
                        );
                        
                     }")
      )
    )
  )
)
```

## Map current yearly and historical average county negative equity rate

We don't have time series data by ZIP Code, but we do have it by county.  Some interesting trends between 2009 and 2019, by comparing two maps.

The first map shows negative equity rates by county in 2019, and the second in 2009.

```{r}

# Filter for 2019
underwater_county_year_2019 <- underwater_county_year %>%
  dplyr::select(state_code, fips_code, state_name, county_name, y2019) %>%
  filter(y2019 >= 4)

# Join zip code coordinates to negative equity by zipcode
underwater_county_year_2019 <- geo_join(counties, underwater_county_year_2019, 'GEOID', 'fips_code', 
how = "inner")  

# Color Scheme
binpal <- colorBin("plasma", underwater_county_year_2019$y2019, 5, pretty = FALSE)

# Draw map
leaflet(underwater_county_year_2019) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%
   #addProviderTiles(providers$Wikimedia) %>%
   addPolygons(fillColor = ~binpal(y2019), weight = 1, smoothFactor = 0.5, opacity = 0.1, fillOpacity = 0.5, color="black", popup = popupTable(underwater_county_year_2019)) %>%
   setView(-95, 39.335359608681216, 4) %>%
   addHomeButton(extent(-98.39355468750001, -89.60449218750001, 39.85072092501597, 43.8899753738369), "IA") %>%
   addHomeButton(extent(-115.44021606445314, -114.89089965820314, 36.04521273039952, 36.318998009207924), "NV") %>%
   addHomeButton(extent(-127.24365234375001, -109.66552734375001, 31.98944183792288, 40.713955826286046), "CA") %>%
   addHomeButton(extent(-88.33145141601564, -87.23281860351564, 41.466399253078876, 41.97276436226528), "IL") %>%
   addHomeButton(extent(-80.82092285156251, -79.72229003906251, 25.517657429994035, 26.12831612064242), "FL") %>%
   addHomeButton(extent(-73.95446777343751, -71.75720214843751, 41.03793062246529, 42.05337156043361), "CT") %>%
   addHomeButton(extent(-74.46807861328126, -73.91876220703126, 40.57484790030712, 40.83199550584334), "NJ2") %>%
   addHomeButton(extent(-75.86883544921876, -73.67156982421876, 39.16414104768742, 40.20824570152502), "NJ1") %>%
   addHomeButton(extent(-76.95098876953126, -74.75372314453126, 38.10214399750345, 39.16201148082406), "MD2") %>%
   addHomeButton(extent(-77.1947479248047, -76.6454315185547, 38.791021386961596, 39.0549177529185), "MD1") %>%
   addHomeButton(extent(-131.74804687500003, -61.43554687500001, 18.812717856407776, 52.908902047770255 ), "U.S.") %>%
   addLogo("https://jeroenooms.github.io/images/banana.gif",
        position = "bottomleft",
        offset.x = 5,
        offset.y = 40,
        width = 100,
        height = 100) %>%
   addLegend("bottomleft", 
             pal = binpal, 
             values = underwater_county_year_2019$y2019,
    title = "% Homes Negative Equity 2019",
    labFormat = labelFormat(prefix = ""),
    opacity = 1
  ) %>%
  addEasyButton(easyButton(
    states = list(
      easyButtonState(
        stateName="unfrozen-markers",
        icon="ion-toggle",
        title="Get Bounding box",
        onClick = JS("
                     function(btn, map) {
                        alert(
                         map.getBounds().getWest() + ', ' + map.getBounds().getEast() + ', ' + map.getBounds().getSouth() + ', ' + map.getBounds().getNorth() 
                        );
                        
                     }")
      )
    )
  )
)

```

## Map 2009 yearly average county negative equity rate

```{r}

# Filter for 2009
underwater_county_year_2009 <- underwater_county_year %>%
  dplyr::select(state_code, fips_code, state_name, county_name, y2009) %>%
  filter(y2009 >= 10)

# Join zip code coordinates to negative equity by zipcode
underwater_county_year_2009 <- geo_join(counties, underwater_county_year_2009, 'GEOID', 'fips_code', 
how = "inner")  

# Color Scheme
binpal <- colorBin("plasma", underwater_county_year_2009$y2009, 5, pretty = FALSE)

# Draw map
leaflet(underwater_county_year_2009) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%
   #addProviderTiles(providers$Wikimedia) %>%
   addPolygons(fillColor = ~binpal(y2009), weight = 1, smoothFactor = 0.5, opacity = 0.1, fillOpacity = 0.5, color="black", popup = popupTable(underwater_county_year_2009)) %>%
   setView(-95, 39.335359608681216, 4) %>%
   addHomeButton(extent(-89.40673828125, -80.61767578125001, 31.23159167205059, 35.755428369259626), "GA") %>%
   addHomeButton(extent(-78.93676757812501, -74.54223632812501, 37.74031329210266, 39.854937988531276), "MD") %>%
   addHomeButton(extent(-76.85485839843751, -72.46032714843751, 38.90385833966778, 40.9840449469281), "NJ") %>%
   addHomeButton(extent(-75.88256835937501, -67.09350585937501, 40.44694705960048, 44.449467536006935), "NE") %>%
   addHomeButton(extent(-90.4888916015625, -86.0943603515625, 40.65147128144057, 42.67839711889055), "IL") %>%
   addHomeButton(extent(-84.79797363281251, -80.40344238281251, 40.91766362458114, 42.93631775765237), "MI") %>%
   addHomeButton(extent(-109.24804687500001, -100.458984375, 37.474858084971046, 41.65649719441145), "CO") %>%
   addHomeButton(extent(-136.45019531250003, -101.29394531250001, 28.38173504322311, 45.583289756006316), "CA") %>%
   addHomeButton(extent(-131.74804687500003, -61.43554687500001, 18.812717856407776, 52.908902047770255 ), "U.S.") %>%
   addLogo("https://jeroenooms.github.io/images/banana.gif",
        position = "bottomleft",
        offset.x = 5,
        offset.y = 40,
        width = 100,
        height = 100) %>%
   addLegend("bottomleft", 
             pal = binpal, 
             values = underwater_county_year_2009$y2009,
    title = "% Homes Negative Equity 2009",
    labFormat = labelFormat(prefix = ""),
    opacity = 1
  ) %>%
  addEasyButton(easyButton(
    states = list(
      easyButtonState(
        stateName="unfrozen-markers",
        icon="ion-toggle",
        title="Get Bounding box",
        onClick = JS("
                     function(btn, map) {
                        alert(
                         map.getBounds().getWest() + ', ' + map.getBounds().getEast() + ', ' + map.getBounds().getSouth() + ', ' + map.getBounds().getNorth() 
                        );
                        
                     }")
      )
    )
  )
)

```


## Explaining the differences
In general, amongst the highest negative equity zips, there's a general trend.  The higher the rate of af american and hispanics, the higher the rate.

```{r}

### Need to add in home price appreciation and mortgage origination as percentage of application
# Pull in NYT classification of urban rural

underwater_zips_2019_join <- underwater_zips_2019 %>%
  inner_join(acs_data_zcta, by=c("zip_code" = "geoid")) %>%
  inner_join(Zip_ZHVI_summary_current_month, by=c("zip_code" = "RegionName")) %>% 
  #%>%
  #filter(!is.na(percent_negative_equity),
   #      pct_white != "NaN",
   #      !is.na(median_household_income)) %>%
  filter(percent_negative_equity >=5) 
 
corr_table_zcta <- underwater_zips_2019_join %>%
  dplyr::select(-matches("code|name|Date|Region|State|Metro|County|City|Month|Quarter|Last|Time")) %>%
  dplyr::select(matches("percent|pct|year")) %>%
  correlate() %>%
  dplyr::select(rowname, percent_negative_equity)

print(corr_table_zcta)
#https://www.dataquest.io/blog/statistical-learning-for-predictive-modeling-r/
#install.packages("moderndive")

score_model <- lm(percent_negative_equity ~ pct_white+pct_black+pct_hispanic+PctFallFromPeak, data = underwater_zips_2019_join)
get_regression_table(score_model) %>% datatable()
get_regression_summaries(score_model) %>% datatable()

ggplot(underwater_zips_2019_join) +
  geom_point(aes(percent_negative_equity, pct_white)) +
  labs(x="Percent Negative Equity", y="Percent White", title="", caption = "") +
  geom_smooth(aes(percent_negative_equity, pct_white), method = "lm", se = FALSE) 

ggplot(underwater_zips_2019_join) +
  geom_point(aes(percent_negative_equity, pct_black)) +
  labs(x="Percent Negative Equity", y="Percent Black", title="", caption = "") +
  geom_smooth(aes(percent_negative_equity, pct_black), method = "lm", se = FALSE) 

ggplot(underwater_zips_2019_join) +
  geom_point(aes(percent_negative_equity, pct_hispanic)) +
  labs(x="Percent Negative Equity", y="Percent Hispanic", title="", caption = "") +
  geom_smooth(aes(percent_negative_equity, pct_hispanic), method = "lm", se = FALSE) 



```

At the county level, the effects are a bit more blunted. 




```{r}

# Join county level data by year with census data
underwater_county_year_join <- underwater_county_year %>%
  inner_join(acs_data_county, by=c("fips_code"="geoid")) %>%
  ungroup() %>%
  inner_join(county_summary_forecast, by=c("state" = "State", "county" = "RegionName")) %>%
  inner_join(rural_urban, by=c("fips_code" = "FIPS")) %>%
  mutate(housing_affordability = Zhvi/median_household_income) #%>%
  #filter(Population_2010 > 100000) %>%
  #filter(y2019 >=1)



#years <- c("y2018","y2017","y2016","y2015","y2014","y2013","y2012","y2011","y2010","y2009")

#for (year in years) {

underwater_county_year_join %>%
  dplyr::select(-matches("code|state|name|county|date|region|Metro|County|City|Month|Quarter|Last|Time|Description")) %>%
  #dplyr::select(matches("pct|median")) %>%
  #filter(!is.na(deparse(substitute(year)))) %>%
  correlate() %>%
  #dplyr::select(rowname, year) %>%
  filter(str_detect(rowname, "^y")) %>% 
  datatable(
    extensions = 'FixedColumns',
    options = list(
    pageLength = 50,
    dom = 't',
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 2)
  )
  )

underwater_county_year_join <- underwater_county_year_join %>%
  dplyr::select(-matches("code|state|name|county|date|region|Metro|County|City|Month|Quarter|Last|Time|Description")) 

#install.packages("moderndive")

score_model <- lm(y2019~ ., subset(underwater_county_year_join, select=c( -y2009, -y2010,-y2011,-y2012,-y2013,-y2014,-y2015,-y2016,-y2017,-y2018)))
get_regression_table(score_model) %>% datatable()
get_regression_summaries(score_model) %>% datatable()

# Ratio of home prices to earnings by county? That's a feature to engineer and test
# Findings -- look at relationship between ZHVI and year.  It used to be much more strongly related to ZHVI in the past than it is now! Suggests it's no longer a function of housing prices.  Something else more important at work here. 
# Tends to be a metro area problem.....
# What would happen if we just filtered non rural


```


### Examine Change Over Time

```{r}

underwater_county_year_ranks <- underwater_county_year %>%
  filter(y2019 >= 1) %>%
  ungroup() %>%
  #na.omit() %>%
  mutate_at(vars(starts_with("y")), funs(round(percent_rank(.)*100,0))) %>% 
  rename_at(vars(starts_with("y")), function(x) paste0(x,"_rank"))

underwater_county_year_ranks <- underwater_county_year_ranks %>%
  left_join(underwater_county_year, by=c("state_code", "fips_code", "state_name", "county_name")) %>%
  left_join(acs_data_county, by=c("fips_code" = "geoid"))
  
print(underwater_county_year_ranks)



```

## Zip Code Data
CoreLogic data. Percent of homes with negative equity for all U.S. ZipCodes. Think data is Q1 2019, but not sure. 
```{r0}
# Load and clean data by fixing zip codes to add a leading zero for northeastern zip codes
underwater <- read_xlsx("../data/input_data/corelogic_underwater_homes.xlsx") %>%
  clean_names() %>%
  mutate(zip_code = clean.zipcodes(zip_code)) %>%
  mutate(pct_homes_negative_equity = round(share_of_homes_in_negative_equity*100, 2)) %>%
  select(-share_of_homes_in_negative_equity)

write_csv(underwater, "../data/output_data/underwater_cleaned_z.csv")

# Check for dup zip codes 

underwater %>%
  group_by(zip_code) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

# Load ZCTA geography data 
acs_variable <- load_variables(2017, "acs5", cache = TRUE)

# Define census api key
census_api_key("549950d36c22ff16455fe196bbbd01d63cfbe6cf")

# Get zip code geography data with total population zcta 2017
acs_zcta_total_pop <- get_acs(geography = "county", variables = c("B06012_001"), geometry = TRUE, survey="acs5", year = 2017)

# Join it
zcta_underwater_geography <- acs_zcta_total_pop %>%
  inner_join(underwater, by = c("GEOID" = "zip_code"))

# anti join

zcta_anti <- underwater %>%
  anti_join(acs_zcta_total_pop, by = c("zip_code" = "GEOID"))

zcta_anti_x <- acs_zcta_total_pop %>%
  anti_join(underwater, by = c("GEOID" = "zip_code"), keepall=TRUE)

```

# Examine 




## Test Correlations

```{r}
# Get zip code geography data with total population zcta 2017
#acs_zcta_total_pop <- get_acs(geography = "zcta", variables = c("B06012_001"), geometry = TRUE, survey="acs5", year = 2017)

# Poverty B06012_002	Estimate!!Total!!Below 100 percent of the poverty level
acs_zcta_total_pov <- get_acs(geography = "zcta", variables = c("B06012_002"), geometry = TRUE, survey="acs5", year = 2017, summary_var = "B06012_001")

# Calculate MOE
acs_zcta_total_pov_x <- acs_zcta_total_pov %>%
  mutate(moe_percent = moe/estimate) %>%
  filter(moe_percent != "Inf") %>%
  filter(moe_percent < .2)

acs_zcta_total_pov_x <- acs_zcta_total_pov_x %>%
  left_join(underwater, by = c("GEOID" = "zip_code")) %>%
  mutate(pov_rate = estimate/summary_est)

  

```

## Examine the Data

High concentrations of negative equity in Connecticut, New Jersey and Maryland.

```{r}

underwater %>%
  filter(!is.na(share_of_homes_in_negative_equity)) %>%
  group_by(state_name) %>%
  summarise(mean_pct_underwater = mean(share_of_homes_in_negative_equity)*100) %>%
  arrange(desc(mean_pct_underwater))

```



```{r}
# County level analysis

underwater_x <- underwater %>%
  left_join(zip_county_crosswalk, by=c("zip_code" = "zip")) %>%
  group_by(zip_code) %>%
  summarise(count=n()) %>%
  arrange(desc(count)) %>%
  filter(count > 1)

# Build tables of share of homes in negative equity


underwater_counties <- underwater %>%
  filter(!is.na(share_of_homes_in_negative_equity)) %>%
  group_by(state_name, county_name) %>%
  summarise(mean_pct_underwater = mean(share_of_homes_in_negative_equity)*100) %>%
  arrange(desc(mean_pct_underwater)) %>%
  mutate(NAME = paste0())

# Get Census Data by County
census_api_key("549950d36c22ff16455fe196bbbd01d63cfbe6cf")

acs_county_total_pop <- get_acs(geography = "county", variables = c("B06012_001"), geometry = FALSE)


```

```{r}

# Define census api key
census_api_key("549950d36c22ff16455fe196bbbd01d63cfbe6cf")

# Examine ACS Variables
acs_variable <- load_variables(2017, "acs5", cache = TRUE)
# B07001 GEOGRAPHICAL MOBILITY IN THE PAST YEAR
  # Mobility B07001_017	Estimate!!Total!!Same house 1 year ago
#  B06012_002	 POVERTY STATUS IN THE PAST 12 MONTHS
  # Total Population B06012_001	Estimate!!Total
  # Poverty B06012_002	Estimate!!Total!!Below 100 percent of the poverty level
  # Poverty B06012_003	Estimate!!Total!!100 to 149 percent of the poverty level
  # Poverty B06012_004	Estimate!!Total!!At or above 150 percent of the poverty level

census_api_key("549950d36c22ff16455fe196bbbd01d63cfbe6cf")

acs_county_total_pop <- get_acs(geography = "county", variables = c("B06012_001"), geometry = FALSE)



acs_zip <- get_acs(geography = "zcta", variables = c("B06012_001"), geometry = FALSE)


acs <- acs %>%
  left_join(acs_variable, by = c("variable" = "name"))

head(orange)

orange %>%
  ggplot(aes(fill = estimate)) + 
  geom_sf(color = NA) + 
  coord_sf(crs = 26911) + 
  scale_fill_viridis_c(option = "magma") 


zip_county_crosswalk <- read_xlsx("../data/input_data/ZIP_COUNTY_092019.xlsx") %>%
  select(zip, county)

```
